---
title: "00_prepare_data"
author: "Jeremias N. Brand"
date: "2022-10-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source('./R/functions.R')
library(ggplot2)
library(readxl)
library(writexl)
library(dplyr)
library(ggrepel)
library(forcats)
library(ape)
library(phytools)
library(phangorn)
library(RColorBrewer)
# library(devtools)
# install_github("glmgen/genlasso")
# install_github("khabbazian/l1ou")
library(l1ou)

BS_iterations = 100
cores = 5
bootstrap = FALSE
```

## Getting the data

The tree and alignments come from this paper: Brand, J. N., G. Viktorin, R. A. W. Wiberg, C. Beisel, and L. Schärer. 2022. Large-scale phylogenomics of the genus Macrostomum (Platyhelminthes) reveals cryptic diversity and novel sexual traits. Molecular Phylogenetics and Evolution 166:107296. [DOI](https://doi.org/10.1016/j.ympev.2021.107296)

Download and extract *noPrimer_H.tar.gz*, *28S_C.aln*, and *trees.tar.gz* from the Zenodo archive at [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.4543289.svg)](https://doi.org/10.5281/zenodo.4543289) and put them into the lib directory.

Get the supplementary table with assignment of 28S here: [mmc6](https://ars.els-cdn.com/content/image/1-s2.0-S1055790321002293-mmc6.xlsx)


Loading morphological data.

```{r}
df <- read_excel("lib/Morphology_data_2021-02-22.xlsx")

df2 <- df |>
  select(Short, ID, total_sprmLen_um:brstlLen_um_per_sperm)

sum_df <- df2 |>
  select(-ID) |>
  group_by(Short) |>
  summarise_all(mean, na.rm = TRUE) |>
  # only those that have ansperm measures
  filter(!is.na(antSpermLength_um))


out_sum_df <- df2 |>
  select(-ID) |>
  group_by(Short) |>
  summarise(
            across(everything(), 
                   list(mean = function(.x) mean(.x, na.rm = TRUE),
                        SD = function(.x) sd(.x, na.rm = TRUE),
                        min =function(.x) min(.x, na.rm = TRUE),
                        max = function(.x) max(.x, na.rm = TRUE),
                        N = function(.x) sum(!is.na(.x))),
                        .names = "{.col}_{.fn}")) |>
  filter(!is.na(antSpermLength_um_mean))

N_df <- df2 |>
  select(-ID) |>
  group_by(Short) |>
  summarise_all(function(x) sum(!is.na(x))) |>
  # only those that have ansperm measures
  filter(!is.na(antSpermLength_um))
```

# Get the phylogenetic tree and change the tip labels to conform to the ones from the morphology table

```{r}
t_UM <- read.tree("lib/C-IQ-TREE-noPrimer-UM.tre")
t_UM$tip.label

# subset to only one representative specimen per species
subset_df <- read_excel('lib/MPE_Tab_S2.xlsx', sheet = 2) |> 
  filter(included_in_comparative_analysis == 'no')
t_UM$tip.label %in% subset_df$Assembly


to_drop <- t_UM$tip.label[t_UM$tip.label %in% subset_df$Assembly]
to_drop
t_UM <- drop.tip(t_UM, to_drop)

sort(t_UM$tip.label)
tree_name = t_UM$tip.label
simple_tips <- gsub('_$','',gsub('([^_]+_?[a-z]?).+', '\\1', t_UM$tip.label))

name_df = data.frame(tree_name, simple_tips)
```

# Subset data

```{r}
EL3 <- read_excel('lib/Evolution_Letters_Tab_S3.xlsx', sheet = 2)

reciprocal_species <- EL3 |>
  filter(inferred_syndrome == 'reciprocal') |>
  pull(Short)

selection <- name_df |>
  filter(simple_tips %in% reciprocal_species) |>
  # some duplicates are still there because of renaming
  filter(!tree_name %in% c(
    "Mac044_MTP_LS_2156",'Maccli_2930_20180705',
    'Maclig_MTP_LS_244','Macqui_KY652218')) |>
  # remove Mac082 because it does not have bristels 
  filter(simple_tips != 'Mac082') |>
  # filter those that have measurments
  filter(simple_tips %in% sum_df$Short)
  
# subset tree to reciprocal species with measurements 
to_drop <- t_UM$tip.label[!t_UM$tip.label %in% selection$tree_name]
to_drop
tre <- drop.tip(t_UM, to_drop)
match(tre$tip.label, selection$tree_name)
tre$tip.label <- selection$simple_tips


df3 <- filter(sum_df, Short %in% selection$simple_tips) |>
  mutate(sperm_ratio = antSpermLength_um/total_sprmLen_um)

per_sp_sum <- df3 |>
  select(-Short,
         total_sprmLen_um,
         antSpermLength_um,
         brstlLen_um_per_sperm,
         sperm_ratio) |>
  summarise(N = n(), 
            across(everything(), 
                   list(mean = function(.x) mean(.x, na.rm = TRUE),
                        SD = function(.x) sd(.x, na.rm = TRUE),
                        min =function(.x) min(.x, na.rm = TRUE),
                        max = function(.x) max(.x, na.rm = TRUE),
                        N = function(.x) sum(!is.na(.x))),
                        .names = "{.col}_{.fn}")) 

# output xlsx
write_xlsx(
  x = list(
    trait_data = df3,
    morphology = filter(out_sum_df, Short %in% df3$Short),
    N = filter(N_df, Short %in% df3$Short),
    across_species_summary = per_sp_sum
  ),
  path = 'output/trait_summary.xlsx'
)

```

Log transform the data

```{r}
df3 <- select(df3, Short, total_sprmLen_um, antSpermLength_um, brstlLen_um_per_sperm, sperm_ratio)
df3[,c(2:4)] <- log10(df3[,c(2:4)])
```

# adjust tree

```{r}
is.ultrametric(tre)
# this is a precision issue
# see http://blog.phytools.org/2017/03/forceultrametric-method-for-ultrametric.html
tre_UM <- ladderize(force.ultrametric(tre))
```

## output tree and trait data

```{r}
comp_data <- list(tre_UM, df3)
saveRDS(comp_data, 'output/comp_data_main.rds')
```

# Now prepare data for the sperm lenght only

Loading morphological data. Only select sperm length.

```{r}
df <- read_excel("lib/Morphology_data_2021-02-22.xlsx")
df2 <- df |>
  select(Short, ID, total_sprmLen_um)

sum_df <- df2 |>
  select(-ID) |>
  group_by(Short) |>
  summarise_all(mean, na.rm = TRUE) 

out_sum_df <- df2 |>
  select(-ID) |>
  group_by(Short) |>
  summarise(
            across(everything(), 
                   list(mean = function(.x) mean(.x, na.rm = TRUE),
                        SD = function(.x) sd(.x, na.rm = TRUE),
                        min =function(.x) min(.x, na.rm = TRUE),
                        max = function(.x) max(.x, na.rm = TRUE),
                        N = function(.x) sum(!is.na(.x))),
                        .names = "{.col}_{.fn}")) 

N_df <- df2 |>
  select(-ID) |>
  group_by(Short) |>
  summarise_all(function(x) sum(!is.na(x))) 
```

# Get the phylogenetic tree and change the tip labels to conform to the ones from the morphology table

```{r}
t_UM <- read.tree("lib/C-IQ-TREE-noPrimer-UM.tre")
t_UM$tip.label

# subset to only one representative specimen per species
subset_df <- read_excel('lib/MPE_Tab_S2.xlsx', sheet = 2) |> 
  filter(included_in_comparative_analysis == 'no')
t_UM$tip.label %in% subset_df$Assembly


to_drop <- t_UM$tip.label[t_UM$tip.label %in% subset_df$Assembly]
to_drop
t_UM <- drop.tip(t_UM, to_drop)

sort(t_UM$tip.label)
tree_name = t_UM$tip.label
simple_tips <- gsub('_$','',gsub('([^_]+_?[a-z]?).+', '\\1', t_UM$tip.label))

name_df = data.frame(tree_name, simple_tips)
```

# Subset data

```{r}
EL3 <- read_excel('lib/Evolution_Letters_Tab_S3.xlsx', sheet = 2)

df3 <- filter(sum_df, Short %in% name_df$simple_tips) |> 
  # remove species with no measurements
  filter(!is.na(total_sprmLen_um))


selection <- name_df |>
  # some duplicates are still there because of renaming
  filter(!tree_name %in% c(
    "Mac044_MTP_LS_2156",'Maccli_2930_20180705',
    'Maclig_MTP_LS_244','Macqui_KY652218')) |>
  # filter those that have measurments
  filter(simple_tips %in% df3$Short)
  
# subset tree to reciprocal species with measurements 
to_drop <- t_UM$tip.label[!t_UM$tip.label %in% selection$tree_name]
to_drop
tre <- drop.tip(t_UM, to_drop)
match(tre$tip.label, selection$tree_name)
tre$tip.label <- selection$simple_tips


per_sp_sum <- df3 |>
  select(-Short,
         total_sprmLen_um) |>
  summarise(N = n(), 
            across(everything(), 
                   list(mean = function(.x) mean(.x, na.rm = TRUE),
                        SD = function(.x) sd(.x, na.rm = TRUE),
                        min =function(.x) min(.x, na.rm = TRUE),
                        max = function(.x) max(.x, na.rm = TRUE),
                        N = function(.x) sum(!is.na(.x))),
                        .names = "{.col}_{.fn}")) 

# output xlsx
write_xlsx(
  x = list(
    trait_data = df3,
    morphology = filter(out_sum_df, Short %in% df3$Short),
    N = filter(N_df, Short %in% df3$Short),
    across_species_summary = per_sp_sum
  ),
  path = 'output/sprmLen_only_trait_summary.xlsx'
)

```

Log transform the data

```{r}
df3[,2] <- log10(df3[,2])
```

# adjust tree

## We want to redo the previous analysis but exclude the hypodermic clade

```{r}
r_clade <- extract.clade(tre, 117, root.edge = 0, collapse.singles = TRUE,
              interactive = FALSE)

plot(r_clade)
tre <- r_clade
```

## adjust the morphological data to match

```{r}
df3 <- df3 |> 
  filter(Short %in% tre$tip.label)
```

```{r}
is.ultrametric(tre)
# this is a precision issue
# see http://blog.phytools.org/2017/03/forceultrametric-method-for-ultrametric.html
tre_UM <- ladderize(force.ultrametric(tre))
```

## output tree and trait data

```{r}
comp_data <- list(tre_UM, df3)
saveRDS(comp_data, 'output/comp_data_sprmLen_only.rds')
```